---
title: "Bring the Heat"
author:
  name: Amanda Dobbyn
  email: aedobbyn@uchicago.edu
output: html_document
---

***

Outline
=====
* About:
    + Info on the [Wildfire](http://theaudl.com/teams/wildfire)
    + Data from [UltiAnalytics](http://www.ultianalytics.com/app/#/5671536392404992/players)
    + More extensive code in `bh_prep.R`, `bh_stats.R`, and `bh_plot.R`
* Pre-process:
    + Standardize names and dates
    + Get data in tidy format
* Analyze:
    + Calculate each player's plus-minus
* Plot:
    + Plot season blocks vs. season goals to determine Wildfire MVP

***

## Munge

__Import packages__  
using `pacman`

```{r import_pacman, echo=TRUE, warning=FALSE, message=FALSE}
library(pacman)
```


```{r load_packages}
p_load(tibble,
       lubridate,
       tidyr,
       dplyr,
       stringr,
       readr,
       data.table,
       ggplot2,
       ggrepel)
```

```{r set_wd, echo=FALSE}
# ------- set wd ---------

setwd("/Users/amanda/Desktop/post_exCog/bring_heat")
```

***

__Import data__  
using `readr`

```{r import_dat, message=FALSE, warning=FALSE}
bring_heat_2 <- read_csv('./ChicagoWildfire2015-stats.csv')
```

Make the data into a [tibble](https://github.com/hadley/tibble)
```{r make_tibble}
bh <- as_tibble(bring_heat_2)
```

Take a look at what we have
```{r}
bh
```
^ yikes

<br /> <br /> <br />

**Manipulate variable names**  
Check out column names
```{r get_names}
names(bh)
```

Get rid of players 8 through 27 because they will never exist  
(There are only ever 7 players per team on the field at once)

```{r remove_phantom_players}
bh <- bh %>%
  select(
    -(`Player 8`:`Player 27`)
  )
```

Replace all spaces in column names with underscores
```{r space_to_underscore}
names(bh) <- str_replace_all(names(bh), " ", "_")
```

Take out "- End of Point" from `Their Score - End of Point` and `Our Score - End of Point`
```{r remove_end_of_point_garbage}
names(bh) <- str_replace_all(names(bh), '_-_End_of_Point', '')
```

Take out the seconds unit at the end of hang time and elapsed time
```{r}
names(bh) <- gsub('_(secs)', '', names(bh), fixed = TRUE) # a bit easier with gsub than with str_replace because of the parentheses around secs
```

Make all variable names lowercase
```{r names_to_lower}
names(bh) <- tolower(names(bh))
```

Rename date/time column to date_time
```{r rename_date_time}
bh <- bh %>% 
  rename(
    date_time = `date/time`
  )
```

Now our column names are more manageable: <br />
```{r new_names}
names(bh)
```


<br /><br /><br />

**Manipulate variable types**  
Make our categorical variables into factors
```{r make_factor}
# make vector of columns we want as factor
want.as.factor <- c('tournamemnt', 'opponent', 'line', 'event_type', 'action', 
                    'passer', 'receiver', 'defender',
                    'player_0', 'player_1', 'player_2', 'player_3', 'player_4', 'player_5',
                    'player_6', 'player_7')


# make these columns factors
bh[, want.as.factor] <- data.frame(apply(bh[, want.as.factor], 2, as.factor))
```

Make our numeric variables numeric
```{r make_numeric}
# make vector of columns we want as numeric
want.as.numeric <- c('point_elapsed_seconds', 'our_score', 'their_score', 
                     'hang_time', 'elapsed_time')

# make these columns numeric
bh[, want.as.numeric] <- data.frame(apply(bh[, want.as.numeric], 2, as.numeric))
```

Check that tournament is always the same and take it out
```{r delete_tournament}
summary(bh$tournamemnt) # careful of the typo -- variable name is tournamenmt

# it's always NA so take it out
bh <- bh %>%
  select(
    -tournamemnt
  )
```

__Clean up the ```opponent``` column__

Check the levels in the `opponent` column
```{r opponent_levels}
levels(bh$opponent)
```

There are multiple columns for the same opponent

Take "vs." in all its forms out of the `opponents` column
```{r remove_vs}
levels(bh$opponent) <- levels(bh$opponent) %>%
                                 str_replace_all(c('vs ', 'vs. ', 'Vs. '), '')

# check that our levels of opponent look right
levels(bh$opponent)

# we have two types of alleycats: Alleycats and AlleyCats. make them into one
bh$opponent[bh$opponent=='AlleyCats'] <- 'Alleycats'
 
```


Check that our levels of opponent look right
```{r checklevels_of_opponent}
levels(bh$opponent)
```

We have two types of Alleycats: `Alleycats` and `AlleyCats`. Combine these levels.
```{r alleycats_levels}
bh$opponent[bh$opponent=='AlleyCats'] <- 'Alleycats'
```

And drop the unused level
```{r drop_unused_level}
bh <- bh %>%
  droplevels()
```

Keep all columns in a `bh_all` tibble so we can safely pare down `bh`
```{r }
bh_all <- bh
```

***

## Tidy

Get game numbers:  
* Individual games (total of 13 this season)
* Game number against each opponent

<br />
Split `date_time` into `date` and `time` columns so that we can differentiate different games played against the same opponent. 
<br />
(These will have the same `opponent` name an different `date`s.)
```{r split_datetime}
bh_all <- separate(bh_all, date_time, into = c("date", "time"), sep=" ") 
```

Get `date` in date format
```{r}
bh_all$date <- ymd(bh_all$date)

# check that the class is date
class(bh_all$date)


```
The `time` column is the same for each date (gives us no 
more information than `date`), so we can safely ignore / get rid of it

***
#### Game numbers

Make a column for each **game number against each opponent**

```{r indiv_gme}
bh_all <- bh_all %>%
  group_by(opponent) %>%
  mutate(
    game = as.numeric(factor(date))
  )

bh_all$game <- factor(bh_all$game) # make game a factor
```

Make a column for each **individual game**

```{r ind.gme}
library(data.table)
bh_all <- setDT(bh_all)[, ind.gme := .GRP, by = .(date, opponent), ]
```

Turn `bh_all` back into tibble
```{r back_to_tibble}
bh_all <- as_tibble(bh_all)
```

Check that our last `ind.gme` value is 13
```{r}
tail(bh_all[, c(1, 3, 23:24)])
```

***

Pare `bh` down to columns we really care about
```{r pare_down}
bh <- bh_all

bh <- bh %>%
  select (
    opponent, our_score, their_score,
    action, 
    passer, receiver, defender, 
    elapsed_time, game, ind.gme
  )
```

__Gather player into one column__  

Combine `passer`, `reciever`, and `defender` into a new column, `actor`.  
Make a `name` column corresponding to name each row in `actor`.
```{r player_column}
bh <- bh %>% 
  tidyr::gather(
    key = actor,
    value = name,
    passer:defender,
    na.rm=T
  )

head(bh[, 4:9])
```

Make `actor` and `name` factors
```{r name_to_factor}
bh$actor <- factor(bh$actor)
bh$name <- factor(bh$name)
```

***

## Get summary stats

Calculate plus minus per `name` (i.e., player) calculated as

$$ plus minus = (goals + blocks) - (throwaways + drops) $$

Arrange `name`s in descending order.
```{r plus_minus}
plus_minus <- bh %>%
  select(
    name, action
  ) %>%
  group_by(
    name
  ) %>%
  summarise(
    goals = sum(action == 'Goal'),
    Ds = sum(action == 'D'),
    throwaways = sum(action == 'Throwaway'),
    drops = sum(action == 'Drop'),
    p_m = (goals + Ds) - (throwaways + drops)
  ) %>%
  arrange(desc(
    p_m
  ))
```

Take out the player "Anonymous" [he is probably a scary hacker]
```{r take_out_anonymous}
mvplus_minus <- plus_minus %>% 
  filter(name != "Anonymous") 
```

Print the table with players' summary stats.  
`p_m` is each player's plus minus.
```{r}
mvplus_minus %>% print(n = 50)
```

***
Plot
=====
<br />

A plot approximating the Wildfire MVP by plotting season Ds vs. season goals.

```{r mvplot}
mvp_plot_2 <- ggplot(data = mvplus_minus, aes(x = Ds, y = goals, 
                                   label = name, angle = 45))
mvp_plot_2 + geom_point() +
  ggtitle("MVP") +
  xlab("Season Blocks") + 
  ylab("Season Goals") +
  geom_text_repel(aes(label = name), 
                  box.padding = unit(0.45, "lines"))
```

<br />

(Of course, this plot doesn't consider the bad things [drops and throwaways], 
so the `p_m` column of the `mvplus_minus` tibble is probably a more accurate
indicator of a player's contributions.)


<br /><br /><br />


